---
title: "Analyzing passerine response to restoration on Lagoon Island, UCSB"
author: "Madeline Berger, Renee Albretch, Robert Heim"
date: "2/17/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(reshape2)
library(paletteer)
library(janitor)
library(lubridate)
library(mgcv)
library(broom)
library(patchwork)
library(RColorBrewer)
library(gt)


#read in CCBER data
birds_all <- read_csv("UCSB_CampusLagoon_BirdSurveyData.csv") %>% 
  clean_names() %>% 
  select(species, lagoon_island_zone_summary, day:year)

#read in breeding bird survey data
bbs_all <- read_csv("bbs_data_all.csv") %>% 
  clean_names()
```

##Introduction: Description of Data and Project

Efforts to remove invasive grasses and restore coastal sage scrub and oak woodlands are widespread throughout California. The goal of most of these restoration projects is to improve and restore ecosystem structure and functioning under the assumption that doing so will support populations of native wildlife (Gardali et al., 2006). Unfortunately, the impact of these restoration projects on local species populations is often unknown (Brawn et al., 2006). This makes it challenging to assess whether a restoration initiative has been successful. Restoration projects benefit from studies that evaluate the impacts of restoration efforts on local wildlife populations because findings from these studies can inform strategies to refine future restoration efforts (Ruiz‐Jaen et al., 2005). Furthermore, providing tangible evidence of restoration success can be beneficial for securing funding for future restoration initiatives. 

The Cheadle Center for Biodiversity and Ecological Restoration (CCBER) has conducted oak woodland and coastal sage scrub restoration activities on Lagoon Island on the campus of the University of California, Santa Barbara intermittently since 2005. Research has suggested that restoration activities in small patches of habitat, such as those on Lagoon Island, can increase reproductive success and breeding density of savannah and woodland birds (Brawn et al., 2006). In this analysis, we evaluate the impact of restoration efforts on Lagoon Island for two species of local songbirds: Song Sparrows and the California Towhee.

###Data cleaning - CCBER
```{r}

birds_all$species <- factor(birds_all$species)

###keeping original data - this is the one you need for modelling. this code just cleans up the month column

birds_all_clean <- birds_all %>% 
  mutate(
    date = paste(day, month, year),
    date = dmy(date)
  ) %>% 
  dplyr::select(-month, -day) %>% 
  mutate(
    month = month(date, label = TRUE)
  ) %>%
  drop_na(month) %>% 
  filter(year != 2017) %>% 
  as.data.frame()

####wrangling data for nice graphs

#break out and then parse dates - this will make graphing nicer
data_long <- melt(birds_all,
        # ID variables - all the variables to keep but not split apart on
    id.vars=c("day", "month","year", "species"),
        # The source columns
    measure.vars=c("lagoon_island_zone_summary"),
        # Name of the destination column that will identify the original
        # column that the measurement came from
    variable.name="location",
    value.name="count"
) %>% 
  mutate(
    species = str_to_lower(species),
    date = paste(day, month, year),
    date = dmy(date)
  ) %>% 
  dplyr::select(-month, -day) %>% 
  mutate(
    month = month(date, label = FALSE)
  ) %>% 
  mutate(species = case_when(
    species %in% "western scrub-jay" ~ "california scrub-jay",
    TRUE ~ species
  ))

#further wrangling if we want each line to represent one bird, also change the name of the scrub jay to match bbs
data_tidy<- uncount(data_long, weights = count) %>% 
  mutate(species = str_to_lower(species)) %>% 
  mutate(species = case_when(
    species %in% "western scrub-jay" ~ "california scrub-jay",
    TRUE ~ species
  ))


#filter by the species we are interested in, and up to 2016 since that is what best matches the bbs 

#tidy version (missing zero counts)
bird_species_tidy <- data_tidy %>% 
  filter(species == "california scrub-jay" | species =="common yellowthroat"| species == "song sparrow" | species == "bewick's wren" | species == "white-crowned sparrow" | species == "house finch" | species == "california towhee" | species == "blue-gray gnatcatcher")

#version with counts
bird_species_count <- data_long %>% 
  filter(species == "california scrub-jay" | species =="common yellowthroat"| species == "song sparrow" | species == "bewick's wren" | species == "white-crowned sparrow" | species == "house finch" | species == "california towhee" | species == "blue-gray gnatcatcher")

```


##Data Cleaning - BBS
```{r}

#change the id codes into species names
bbs_clean <- bbs_all %>% 
  mutate(
    species = case_when(
      aou %in% 5190 ~ "house finch",
      aou %in% 5540 ~ "white-crowned sparrow",
      aou %in% 5810 ~ "song sparrow",
      aou %in% 5911 ~ "california towhee",
      aou %in% 6810 ~ "common yellowthroat",
      aou %in% 7190 ~ "bewick's wren",
      aou %in% 7510 ~ "blue-gray gnatcatcher",
      aou %in% 4812 ~ "california scrub jay"
    )
  ) %>% 
  select(-aou, -country)


```


##Summary graphs - CCBER

```{r}

#create dfs that summarize yearly and monthly totals
bird_summary_year <- bird_species_tidy %>% 
  select(-location) %>%
  #filter(species != "house finch") %>%
  filter(year != 2017) %>% 
  group_by(year, species) %>% 
  summarize(
    count = length(species)
  )

bird_summary_month_yr <- bird_species_tidy %>% 
  select(-location) %>% 
  group_by(month, year, species) %>% 
  summarize(
    count = length(species)
  )

#change to sparrow bewicks only
sparrow_towhee_only <- bird_species_tidy %>% 
  select(-location) %>% 
  filter(species == "song sparrow" | species == "california towhee") %>% 
  group_by(year, species) %>% 
  summarize(
    count = length(species)
  )

#yearly abdunace graph with all the birds
yearly_abundance_graph <- ggplot(bird_summary_year, aes(x = year, y = count, group = species))+ 
  geom_line(aes(color = species))+
  theme_minimal()+
  labs(x = "Year", y = "Count", title = "Yearly passerine count at Lagoon Island 1996 - 2016", color = "Species")+
  scale_y_continuous(expand = c(0,0))+
  geom_vline(xintercept = 2005, linetype = "dashed")+
  scale_color_paletteer_d("ggthemes::Summer")

yearly_abundance_graph
ggsave("count_all_yr.png", width = 6, height = 3, units = "in")

#just sparrow and finch 

sparrow_towhee_gg <- ggplot(sparrow_towhee_only, aes(x = year, y = count, group = species)) +
  geom_line(aes(color = species))+
  theme_minimal()+
  labs(x = "Year", y = "Count", title = "Yearly house finch and song sparrow count, Lagoon Island 1996 - 2017", color = "Species")+
  scale_y_continuous(expand = c(0,0))+
  geom_vline(xintercept = 2005, linetype = "dashed")+
  scale_color_manual(breaks = c("song sparrow", "california towhee"), values = c("paleturquoise2", "seagreen3"))
  
sparrow_towhee_gg


```


##Analysis: Song Sparrow

###CCBER data
**Model**
We fit a GAM on our observation data, and then used that to predict yearly abundance, in July, for the species.
<br>
```{r, results=FALSE}
#go back to data read in `birds_all_clean` - pre major wrangling 
#pick a species

birds_ssparrow <- birds_all_clean %>%
  filter(species == "Song Sparrow")

#run GAM on CCBER data

ccber_ssparrow_model <- mgcv::gam(lagoon_island_zone_summary ~ s(year) + month, family = poisson, data = birds_ssparrow) #this gives you the log transformed abudnance

summary(ccber_ssparrow_model)

#create new data frame to use with the model
newd <- data.frame(year = 1996:2016, month = "Jul")

#predict abundance using new data, type predict shows actual abudance instead of log

ssparrow_predict <- predict.gam(ccber_ssparrow_model, newd, type="response")

ssparrow_predict

#notes /issues
#didn't month because I couldn't figure it out
#how can we ask the models for integers

plot(ccber_ssparrow_model)


ssparrow_test <- lm(lagoon_island_zone_summary~year+ month, data = birds_ssparrow)

#summary(ssparrow_test)
```

Tidy model and visualize 
<br>
```{r}

ccber_ssparrow_tidy <- broom::tidy(ccber_ssparrow_model) #this didn't end up being necessary

#we want the actual counts per year for song sparrows, so use the `birds_summmary_year` df

ssparrow_by_year <- bird_summary_month_yr %>% 
  filter(species == "song sparrow") %>% 
  filter(month == 7) %>%
  filter(year != 2017)

#we need data for 1996 - 2000, going to create a dataframe with count = 0 

month <- rep(7,7)
year <- c(1996, 1997, 1998, 1999, 2000, 2004, 2005)
species <- c("song sparrow", "song sparrow", "song sparrow", "song sparrow","song sparrow","song sparrow", "song sparrow")
count <- rep(0,7)

#make into dataframe
missing_ss <- data.frame(month, year,species,count)

ssparrow_jul <- bind_rows(missing_ss, ssparrow_by_year)

#bind obs and predictions
ssparrow_complete <- data.frame(ssparrow_jul, ssparrow_predict)


ssparrow <- ggplot(data = ssparrow_complete, aes(x = year, y = count)) +
  geom_point()+
  geom_line(aes(x = year, y = ssparrow_predict), color = "paleturquoise3") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance")

ssparrow

#ggsave("count_v_predict.png", width = 6, height = 3, units = "in")
```


**Calculate growth rates**
```{r}

#create a new column in sparrow complete that calculates r based on the year before using the log equation

ssparrow_gr <- ssparrow_complete %>% 
  mutate(
    r = log(lead(ssparrow_predict)/ssparrow_predict)
  )

#graph - what should we test?

ssparrow_r <- ggplot(ssparrow_gr, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ssparrow_r

ggsave("ssparrow_gr_ccber.png", width = 6, height = 3, units = "in")


#find averages - I didn't make a new data frame for each test I just changed the names and re -ran this code

ss_ccber_before <- ssparrow_gr %>% 
  filter(year < 2006) %>% 
  filter(year != 2016)

ss_avg_bef_ccber <- mean(ss_ccber_before$r) #this always stays the same
sd_bef<- sd(ss_ccber_before$r) # feed in a vector
se_bef <- sd_bef/sqrt(nrow(ss_ccber_before)) #calculate se 

ss_ccber_after <- ssparrow_gr %>% 
  filter(year > 2010) %>% 
  filter(year != 2016)

ss_avg_after_5 <- mean(ss_ccber_after$r)
sd_after_5<- sd(ss_ccber_after$r) # feed in a vector
se_after_5<- sd_bef/sqrt(nrow(ss_ccber_after))


```

Combine graphs using patchwork - this is more for knitting
```{r}

ssparrow + ssparrow_r

```


###BBS
**Model**
```{r}
#filter out song sparrow

bbs_ssparrow <- bbs_clean %>% 
  filter(species == "song sparrow") %>% 
  group_by(year) %>% 
  summarize(
    species_total = sum(species_total)
  )

#run GAM 
bbs_ssparrow_model <- mgcv::gam(species_total ~ s(year), family = poisson, data = bbs_ssparrow)

summary(bbs_ssparrow_model)

plot(bbs_ssparrow_model)

#create new data frame for predictions
newd_bbs <- data.frame(year = 1996:2016)

#predict abundance using new data

ssparrow_predict_bbs <- predict.gam(bbs_ssparrow_model, newd_bbs, type="response")

ssparrow_predict_bbs

```

Join observations and predictions
```{r}

ssparrow_complete_bbs <- data.frame(bbs_ssparrow, ssparrow_predict_bbs)

ssparrow_gr_bbs <- ssparrow_complete_bbs %>% 
  mutate(
    r = log(lead(ssparrow_predict_bbs)/ssparrow_predict_bbs)
  )

#note - this graph of the model prediction + observation for BBS is all the data because it is compiled by year
ssparrow_bbs <- ggplot(data = ssparrow_complete_bbs, aes(x = year, y = species_total)) +
  geom_point()+
  geom_line(aes(x = year, y = ssparrow_predict_bbs), color = "steelblue3") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance")

ssparrow_bbs

ggsave("count_v_predict_bbs.png", width = 6, height = 3, units = "in")


ssparrow_bbs_r <- ggplot(ssparrow_gr_bbs, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ssparrow_bbs_r

ggsave("ssparrow_bbs_growthrate.png", width = 6, height = 3, units = "in")

```

**Calculate growth rates**
```{r}
#df
ss_bbs_before <- ssparrow_gr_bbs %>% 
  filter(year < 2006) %>% 
  filter(year != 2016)
#numbers
ss_bbs_bef <- mean(ss_bbs_before$r)
sd_bbs_bef<- sd(ss_bbs_before$r) # feed in a vector
se_bbs_bef <- sd_bef/sqrt(nrow(ss_bbs_before)) #calculate se 

#df
ss_bbs_after <- ssparrow_gr_bbs %>% 
  filter(year > 2010) %>% 
  filter(year != 2016)

#numbers
ss_bbs_after_5 <- mean(ss_bbs_after$r)
sd_bbs_after_5<- sd(ss_bbs_after$r) # feed in a vector
se_bbs_after_5 <- sd_bef/sqrt(nrow(ss_bbs_after))


```

**Test if difference between growth rates for CCBER and BBS is significant**

```{r}

#three tests: before, after 2005, after 2010. 

t.test(ss_ccber_before$r, ss_bbs_before$r, conf.level = 0.95, paired = TRUE)

t.test(ss_ccber_after$r,ss_bbs_after$r, conf.level = 0.95, paired = TRUE)

```



##Analysis: California towhee

###CCBER
**Model**
```{r}
#Go back to data read in `birds_all_clean` - pre major wrangling 
#Pick a species - California towhee

birds_ctowhee <- birds_all_clean %>%
  filter(species == "California Towhee")

#Run GAM on CCBER data

ccber_ctowhee_model <- mgcv::gam(lagoon_island_zone_summary ~ s(year) + month, family = poisson, data = birds_ctowhee) #this gives you the log transformed abudnance

summary(ccber_ctowhee_model)

#create new data frame to use with the model
newd <- data.frame(year = 1996:2016, month = "Jul")

#predict abundance using new data, type predict shows actual abudance instead of log

ctowhee_predict <- predict.gam(ccber_ctowhee_model, newd, type="response")

ctowhee_predict

#notes /issues
#didn't month because I couldn't figure it out
#how can we ask the models for integers

plot(ccber_ctowhee_model)

ctowhee_test <- lm(lagoon_island_zone_summary~year+ month, data = birds_ctowhee)

summary(ctowhee_test)
```

**Tidy model and visuzalize**

```{r}

ccber_ctowhee_tidy <- broom::tidy(ccber_ctowhee_model) #this didn't end up being necessary

#we want the actual counts per year for towhees, so use the `birds_summmary_year` df

ctowhee_by_year <- bird_summary_month_yr %>% 
  filter(species == "california towhee") %>% 
  filter(month == 7) %>%
  filter(year != 2017)

#we need data for 1996 - 2000, going to create a dataframe with count = 0 
#years will be different for Bewick's Wren

month <- rep(6,6)
year <- c(1996, 1997, 1998, 2000, 2001, 2005)
species <- c("california towhee", "california towhee", "california towhee", "california towhee","california towhee","california towhee")
count <- rep(0,6)

#make into dataframe
missing_ct <- data.frame(month, year,species,count)

ctowhee_jul <- bind_rows(missing_ct, ctowhee_by_year)

#bind obs and predictions
ctowhee_complete <- data.frame(ctowhee_jul, ctowhee_predict)


ctowhee <- ggplot(data = ctowhee_complete, aes(x = year, y = count)) +
  geom_point()+
  geom_line(aes(x = year, y = ctowhee_predict), color = "palegreen3") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance", title="Observed California towhee counts versus predicted abundance")

ctowhee

ggsave("count_v_predict.png", width = 6, height = 3, units = "in")
```

**Calculate growth rates**

```{r}

#create a new column in ctowhee complete that calculates r based on the year before using the log equation

ctowhee_gr <- ctowhee_complete %>% 
  mutate(
    r = log(lead(ctowhee_predict)/ctowhee_predict)
  )

#graph - what should we test?

ctowhee_r <- ggplot(ctowhee_gr, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ctowhee_r

ggsave("ctowhee_gr_ccber.png", width = 6, height = 3, units = "in")


#find averages - I didn't make a new data frame for each test I just changed the names and re -ran this code

ct_ccber_before <- ctowhee_gr %>% 
  filter(year < 2006) %>% 
  filter(year != 2016)

ct_avg_bef_ccber <- mean(ct_ccber_before$r) #this always stays the same
sd_bef<- sd(ct_ccber_before$r) # feed in a vector
se_bef <- sd_bef/sqrt(nrow(ct_ccber_before)) #calculate se 

ct_ccber_after <- ctowhee_gr %>% 
  filter(year > 2005) %>% 
  filter(year != 2016)

#redo for 5 and 10 years after restoration

ct_avg_after_5 <- mean(ct_ccber_after$r)
sd_after_5<- sd(ct_ccber_after$r) # feed in a vector
se_after_5<- sd_bef/sqrt(nrow(ct_ccber_after))

ct_ccber_after_10 <- ctowhee_gr %>% 
  filter(year > 2010) %>% 
  filter(year != 2016)

ct_avg_after_10 <- mean(ct_ccber_after_10$r)
sd_after_10<- sd(ct_ccber_after_10$r) # feed in a vector
se_after_10<- sd_bef/sqrt(nrow(ct_ccber_after_10))



```

**Combine graphs using patchwork - this is more for knitting**

```{r}

ctowhee + ctowhee_r

# May want to change y-axis scale

```


###BBS
**Model**
```{r}

#filter out song sparrow

bbs_ctowhee <- bbs_clean %>% 
  filter(species == "california towhee") %>% 
  group_by(year) %>% 
  summarize(
    species_total = sum(species_total)
  )

#run GAM 
bbs_ctowhee_model <- mgcv::gam(species_total ~ s(year), family = poisson, data = bbs_ctowhee)

summary(bbs_ctowhee_model)

plot(bbs_ctowhee_model)

#create new data frame for predictions
newd_ctowhee_bbs <- data.frame(year = 1996:2016)

#predict abundance using new data

ctowhee_predict_bbs <- predict.gam(bbs_ctowhee_model, newd_ctowhee_bbs, type="response")

ctowhee_predict_bbs

```

**Join observations and predictions**
```{r}

ctowhee_complete_bbs <- data.frame(bbs_ctowhee, ctowhee_predict_bbs)

ctowhee_gr_bbs <- ctowhee_complete_bbs %>% 
  mutate(
    r = log(lead(ctowhee_predict_bbs)/ctowhee_predict_bbs)
  )

#note - this graph of the model prediction + observation for BBS is all the data because it is compiled by year
ctowhee_bbs <- ggplot(data = ctowhee_complete_bbs, aes(x = year, y = species_total)) +
  geom_point()+
  geom_line(aes(x = year, y = ctowhee_predict_bbs), color = "springgreen4") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance")

ctowhee_bbs

ggsave("count_v_predict_bbs.png", width = 6, height = 3, units = "in")


ctowhee_bbs_r <- ggplot(ctowhee_gr_bbs, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ctowhee_bbs_r

ggsave("ctowhee_bbs_growthrate.png", width = 6, height = 3, units = "in")

```

**Calculate growth rates**
```{r}

#df
ct_bbs_before <- ctowhee_gr_bbs %>% 
  filter(year < 2006) %>% 
  filter(year != 2016)

#numbers before restoration
ct_bbs_bef <- mean(ct_bbs_before$r)
sd_bbs_bef<- sd(ct_bbs_before$r) # feed in a vector
se_bbs_bef <- sd_bef/sqrt(nrow(ct_bbs_before)) #calculate se 

#df
ct_bbs_after <- ctowhee_gr_bbs %>% 
  filter(year > 2005) %>% 
  filter(year != 2016)

#numbers 5 and 10 years after restoration

ct_bbs_after_5 <- mean(ct_bbs_after$r)
sd_bbs_after_5<- sd(ct_bbs_after$r) # feed in a vector
se_bbs_after_5 <- sd_bef/sqrt(nrow(ct_bbs_after))

ct_bbs_after_10 <- ctowhee_gr_bbs %>% 
  filter(year > 2010) %>% 
  filter(year != 2016)

ct_avg_bbs_after_10 <- mean(ct_bbs_after_10$r)
sd_bbs_after_10<- sd(ct_bbs_after_10$r) # feed in a vector
se_bbs_after_10 <- sd_bef/sqrt(nrow(ct_bbs_after_10))
                               
#Make a dataframe with averages - try again later
# 
# Dates <- c(2005, 2010,2015)
# Before <- c(ss_bbs_bef, ss_bbs_bef, ss_bbs_bef)
# After <- c(ss_bbs_after_5, ss_bbs_after_10, ss_bbs_after_15)
# 
# average_table_bbs <- data.frame(Dates, Before, After)


```


**Test if difference between growth rates for CCBER and BBS is significant for california towhee**

```{r}

#three tests: before, after 2005, after 2010. 

t.test(ct_ccber_before$r, ct_bbs_before$r, conf.level = 0.95, paired= TRUE)

t.test(ct_ccber_after$r,ct_bbs_after$r, conf.level = 0.95, paired = TRUE)

t.test(ct_ccber_after_10$r,ct_bbs_after_10$r, conf.level = 0.95, paired = TRUE)

```


