---
title: "Cyber_bird_data"
author: "Madeline Berger"
date: "2/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(reshape2)
library(paletteer)
library(janitor)
library(lubridate)
library(mgcv)
library(broom)
library(patchwork)
library(RColorBrewer)


#read in our data
birds_all <- read_csv("UCSB_CampusLagoon_BirdSurveyData.csv") %>% 
  clean_names() %>% 
  select(species, lagoon_island_zone_summary, day:year)

#read in breeding bird survey data
bbs_all <- read_csv("bbs_data_all.csv") %>% 
  clean_names()
```

##Introduction: Description of Data and Project

The CCBER at UC Santa Barbara has been monitoring bird populations on and near campus since 1996. In 2005, the University began oak restoration...

###Data cleaning - CCBER
```{r}

birds_all$species <- factor(birds_all$species)

###keeping original data - this is the one you need for modelling

birds_all_clean <- birds_all %>% 
  mutate(
    date = paste(day, month, year),
    date = dmy(date)
  ) %>% 
  dplyr::select(-month, -day) %>% 
  mutate(
    month = month(date, label = TRUE)
  ) %>%
  drop_na(month) %>% 
  filter(year != 2017)

#break out and then parse dates - this will make graphing nicer
data_long <- melt(birds_all,
        # ID variables - all the variables to keep but not split apart on
    id.vars=c("day", "month","year", "species"),
        # The source columns
    measure.vars=c("lagoon_island_zone_summary"),
        # Name of the destination column that will identify the original
        # column that the measurement came from
    variable.name="location",
    value.name="count"
) %>% 
  mutate(
    species = str_to_lower(species),
    date = paste(day, month, year),
    date = dmy(date)
  ) %>% 
  dplyr::select(-month, -day) %>% 
  mutate(
    month = month(date, label = FALSE)
  ) %>% 
  mutate(species = case_when(
    species %in% "western scrub-jay" ~ "california scrub-jay",
    TRUE ~ species
  ))

#if we want each line to represent one bird, also change the name of the scrub jay to match bbs
data_tidy<- uncount(data_long, weights = count) %>% 
  mutate(species = str_to_lower(species)) %>% 
  mutate(species = case_when(
    species %in% "western scrub-jay" ~ "california scrub-jay",
    TRUE ~ species
  ))


#filter by the species we are interested in, and up to 2016 since that is what best matches the bbs 

#tidy version (missing zero counts)
bird_species_tidy <- data_tidy %>% 
  filter(species == "california scrub-jay" | species =="common yellowthroat"| species == "song sparrow" | species == "bewick's wren" | species == "white-crowned sparrow" | species == "house finch" | species == "california towhee" | species == "blue-gray gnatcatcher")

#version with counts
bird_species_count <- data_long %>% 
  filter(species == "california scrub-jay" | species =="common yellowthroat"| species == "song sparrow" | species == "bewick's wren" | species == "white-crowned sparrow" | species == "house finch" | species == "california towhee" | species == "blue-gray gnatcatcher")




```


##Data Cleaning - BBS
```{r}

bbs_clean <- bbs_all %>% 
  mutate(
    species = case_when(
      aou %in% 5190 ~ "house finch",
      aou %in% 5540 ~ "white-crowned sparrow",
      aou %in% 5810 ~ "song sparrow",
      aou %in% 5911 ~ "california towhee",
      aou %in% 6810 ~ "common yellowthroat",
      aou %in% 7190 ~ "bewick's wren",
      aou %in% 7510 ~ "blue-gray gnatcatcher",
      aou %in% 4812 ~ "california scrub jay"
    )
  ) %>% 
  select(-aou, -country)




```




##Summary graphs - CCBER

```{r}

#create dfs that summarize yearly and monthly totals
bird_summary_year <- bird_species_tidy %>% 
  select(-location) %>%
  #filter(species != "house finch") %>%
  filter(year != 2017) %>% 
  group_by(year, species) %>% 
  summarize(
    count = length(species)
  )

bird_summary_month_yr <- bird_species_tidy %>% 
  select(-location) %>% 
  group_by(month, year, species) %>% 
  summarize(
    count = length(species)
  )


finch_sparrow_only <- bird_species_tidy %>% 
  select(-location) %>% 
  filter(species == "song sparrow" | species == "house finch") %>% 
  group_by(year, species) %>% 
  summarize(
    count = length(species)
  )

#graph minus house finch
yearly_abundance_graph <- ggplot(bird_summary_year, aes(x = year, y = count, group = species))+ 
  geom_line(aes(color = species))+
  theme_minimal()+
  labs(x = "Year", y = "Count", title = "Yearly passerine count at  Lagoon Island 1996 - 2017", color = "Species")+
  scale_y_continuous(expand = c(0,0))+
  geom_vline(xintercept = 2005, linetype = "dashed")

yearly_abundance_graph

yearly_abundance_graph +
  scale_color_paletteer_d("ggthemes::Summer")

ggsave("count_all_yr.png", width = 6, height = 3, units = "in")

#just sparrow and finch 

sparrow_finch_gg <- ggplot(finch_sparrow_only, aes(x = year, y = count, group = species)) +
  geom_line(aes(color = species))+
  theme_minimal()+
  labs(x = "Year", y = "Count", title = "Yearly house finch and song sparrow count at  Lagoon Island 1996 - 2017", color = "Species")+
  scale_y_continuous(expand = c(0,0))+
  geom_vline(xintercept = 2005, linetype = "dashed")+
  scale_color_paletteer_d("calecopal::lake")
  
sparrow_finch_gg


#by month - comment out because we don't really want it and it doesn't run 

# monthy_ab_graph <- ggplot(bird_summary_month, aes(x = month, y = count, group = species))+ 
#   geom_line(aes(color = species))+
#   theme_minimal()+
#   labs(x = "Month", y = "Count", title = "Monthly passerine count at Lagoon Island 1996 - 2017")+
#   scale_y_continuous(expand = c(0,0))
# 
# monthy_ab_graph



```

Exporting Graphs
```{r}
png("graph_1.png", units = "in", width = 6, height = 3, res = 300)

ggplot(bird_summary_year, aes(x = year, y = count, group = species))+ 
  geom_line(aes(color = species))+
  theme_minimal()+
  labs(x = "Year", y = "Count", title = "Yearly passerine count at  Lagoon Island 1996 - 2017")+
  scale_y_continuous(expand = c(0,0)) +
  geom_vline(xintercept = 2005)


```


##Analysis

###CCBER data - Song Sparrow
```{r}
#go back to count
#pick a species

birds_ssparrow <- birds_all_clean %>%
  filter(species == "Song Sparrow")

#run GAM on CCBER data

ccber_ssparrow_model <- mgcv::gam(lagoon_island_zone_summary ~ s(year) + month, family = poisson, data = birds_ssparrow) #this gives you the log transformed abudnance

summary(ccber_ssparrow_model)

#create new data frame to use with the model
newd <- data.frame(year = 1996:2016, month = "Jul")

#predict abundance using new data

ssparrow_predict <- predict.gam(ccber_ssparrow_model, newd, type="response")

ssparrow_predict

#notes /issues
#didn't month because I couldn't figure it out
#how can we ask the models for integers

plot(ccber_ssparrow_model)
```

Tidy model and visualize 

```{r}

ccber_ssparrow_tidy <- broom::tidy(ccber_ssparrow_model)

#we want the actual counts per year for song sparrows, so use the `birds_summmary_year` df

ssparrow_by_year <- bird_summary_month_yr %>% 
  filter(species == "song sparrow") %>% 
  filter(month == 7) %>%
  filter(year != 2017)

#we need data for 1996 - 2000, going to create a dataframe with count = 0 

month <- rep(7,7)
year <- c(1996, 1997, 1998, 1999, 2000, 2004, 2005)
species <- c("song sparrow", "song sparrow", "song sparrow", "song sparrow","song sparrow","song sparrow", "song sparrow")
count <- rep(0,7)

#make into dataframe
missing_ss <- data.frame(month, year,species,count)

ssparrow_jul <- bind_rows(missing_ss, ssparrow_by_year)

#bind obs and predictions
ssparrow_complete <- data.frame(ssparrow_jul, ssparrow_predict)

#bind 

ssparrow <- ggplot(data = ssparrow_complete, aes(x = year, y = count)) +
  geom_point()+
  geom_line(aes(x = year, y = ssparrow_predict), color = "paleturquoise3") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance")

ssparrow

ggsave("count_v_predict.png", width = 6, height = 3, units = "in")
```

Calculate growth rates
```{r}

#create a new column in sparrow complete that calculates r based on the year before

ssparrow_gr <- ssparrow_complete %>% 
  mutate(
    r = log(lag(ssparrow_predict)/ssparrow_predict)
  )

#graph - what should we test?

ssparrow_r <- ggplot(ssparrow_gr, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ssparrow_r

ggsave("ssparrow_growthrate.png", width = 6, height = 3, units = "in")


#find averages

ss_ccber_before <- ssparrow_gr %>% 
  filter(year < 2006) %>% 
  filter(year != 1996)

mean(ss_ccber_before$r)


ss_ccber_after <- ssparrow_gr %>% 
  filter(year > 2005)

mean(ss_ccber_after$r)
```








Combine graphs using patchwork - this is more for knitting
```{r}

ssparrow + ssparrow_r

```


###BBS - Song Sparrow
```{r}
#filter out song sparrow

bbs_ssparrow <- bbs_clean %>% 
  filter(species == "song sparrow") %>% 
  group_by(year) %>% 
  summarize(
    species_total = sum(species_total)
  )

#run GAM 
bbs_ssparrow_model <- mgcv::gam(species_total ~ s(year), family = poisson, data = bbs_ssparrow)

summary(bbs_ssparrow_model)

plot(bbs_ssparrow_model)

#create new data frame for predictions
newd_bbs <- data.frame(year = 1996:2016)

#predict abundance using new data

ssparrow_predict_bbs <- predict.gam(bbs_ssparrow_model, newd_bbs, type="response")

ssparrow_predict_bbs

```

Join observations and predictions
```{r}

ssparrow_complete_bbs <- data.frame(bbs_ssparrow, ssparrow_predict_bbs)

ssparrow_gr_bbs <- ssparrow_complete_bbs %>% 
  mutate(
    r = log(lag(ssparrow_predict_bbs)/ssparrow_predict_bbs)
  )

ssparrow_bbs <- ggplot(data = ssparrow_complete_bbs, aes(x = year, y = species_total)) +
  geom_point()+
  geom_line(aes(x = year, y = ssparrow_predict_bbs), color = "steelblue3") +
  theme_minimal()+
  labs(x = "Year", y = "Abundance")

ssparrow_bbs

ggsave("count_v_predict_bbs.png", width = 6, height = 3, units = "in")


ssparrow_bbs_r <- ggplot(ssparrow_gr_bbs, aes(x = year, y = r))+
  geom_line(color = "sienna2")+
  theme_minimal()+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 1995)+
  geom_vline(xintercept = 2005, linetype = "dotted")+
  labs(x = "Year", y = "Growth rate (r)")

ssparrow_bbs_r

ggsave("ssparrow_bbs_growthrate.png", width = 6, height = 3, units = "in")

```


